---
title: 'Task 2: Visualize and Analyze Energy Data'
author: "Esteban Villalobos Gómez"
date: "`r format(Sys.Date(), '%m %d, %Y')`"
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: hide
subtitle: 'Xtol: Deep Analytics and Visualization 2017.3'
---
# The "Individual household electric power consumption" Data Set
```{r include=FALSE}
library(RMySQL)
library(dplyr)
library(ggplot2)
library(lubridate)
library(plotly)
```
The dataset contains data from three power sub-meters installed in a house near central France, each measures specific areas of the house:

 - Sub_metering_1: measures a dishwasher, an oven and a microwave
 - Sub_metering_2: measures the washing-machine, a tumble-drier, a refrigerator and a light
 - Sub_metering_3: measures a water-heater and an air-conditioner
```{r include=FALSE}
con = dbConnect(
  MySQL(),
  user = 'deepAnalytics',
  password = 'Sqltask1234!',
  dbname = 'dataanalytics2018',
  host = 'data-analytics-2018.cbrosir2cswx.us-east-1.rds.amazonaws.com'
)

## List the tables contained in the database
#dbListTables(con)

## Use asterisk to specify all attributes for download
power_2007 <- dbGetQuery(con, "SELECT * FROM yr_2007")
power_2008 <- dbGetQuery(con, "SELECT * FROM yr_2008")
power_2009 <- dbGetQuery(con, "SELECT * FROM yr_2009")
```

## Initial Data visualization

The original data set comes with Date and Time columns (both as string), lets convert them into a single field called DateTime:
```{r}
# Merge datasets
power_df <-
  bind_rows(power_2007, power_2008, power_2009)

# Create DateTime Column
## Combine Date and Time attribute values in a new attribute column
power_df <-cbind(power_df,paste(power_df$Date,power_df$Time), stringsAsFactors=FALSE)

## Give the new attribute in the 6th column a header name 
## NOTE: if you downloaded more than 5 attributes you will need to change the column number)
# str(power_df)
colnames(power_df)[11] <-"DateTime"
# Move the DateTime column to be the first column (for convenience)
power_df <- power_df[,c(ncol(power_df), 1:(ncol(power_df)-1))]
power_df
```
The DateTime column is still a String, in order to use it's time components needs to be converted into a POSIXct date type.
```{r}
## Convert DateTime from character to POSIXct 
power_df$DateTime <- as.POSIXct(power_df$DateTime, tz="Europe/Paris", format="%Y-%m-%d %H:%M:%S")
## Add the time zone
attr(power_df$DateTime, "tzone") <- "Europe/Paris"
power_df
```

Now, it's possible to decompose the DateTime field into separate components for further analysis later on.
```{r}
## Create "year" attribute with lubridate
power_df$year <- year(power_df$DateTime)
power_df$quarter <- quarter(power_df$DateTime)
power_df$month <- month(power_df$DateTime)
power_df$week <- week(power_df$DateTime)
power_df$weekday <- wday(power_df$DateTime, week_start=1)
power_df$day <- day(power_df$DateTime)
power_df$hour <- hour(power_df$DateTime)
power_df$minute <- minute(power_df$DateTime)

power_df
```

Let's try to visualize a single week:
```{r}
## Subset the second week of 2008 - All Observations
houseWeek <- filter(power_df, year == 2008 & week == 2)
## Plot subset houseWeek
plot(houseWeek$Sub_metering_1)
```

There are too many data points displayed, therefore is difficult to infer anything from it, let's zoom in into a single day:
```{r}
# Visualize a Single Day with Plotly
## Subset the 9th day of January 2008 - All observations
houseDay <- filter(power_df, year == 2008 & month == 1 & day == 9)
## Plot sub-meter 1
plot_ly(houseDay, x = ~houseDay$Time, y = ~houseDay$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 layout(title = "Power Consumption January 9Th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```
The visualization is better, so let's add the three sub-meters to the graph:
```{r}
## Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations 
plot_ly(houseDay, x = ~houseDay$DateTime, y = ~houseDay$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseDay$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~houseDay$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9Th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```
We can reduce the granularity from *one observation per minute* to *one observation every 10 minutes* to get a better plot, this will be called a **Day** visualization:
```{r}
## Subset the 9th day of January 2008 - 10 Minute frequency
houseDay10 <- filter(power_df, year == 2008 & month == 1 & day == 9 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))

## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10 Minute frequency
plot_ly(houseDay10, x = ~houseDay10$DateTime, y = ~houseDay10$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseDay10$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~houseDay10$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption January 9Th, 2008",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```
Now let's look at another time period: the 3rd week of April 2009 - 10 Minute frequency
```{r}
## Subset the 3rd week of April 2009 - 10 Minute frequency
houseWeek16 <- power_df %>%
  filter(year == 2009 & month == 4 & week == 16 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))

## Plot sub-meter 1, 2 and 3 with title, legend and labels - 10 Minute frequency
plot_ly(houseWeek16, x = ~houseWeek16$DateTime, y = ~houseWeek16$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption April 16Th-22Nd, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))

```
Too many samples, let's reduce them to one per hour:
```{r}
## Subset the 3rd week of April 2009 - 60 Minute frequency
houseWeek16 <- power_df %>%
  filter(year == 2009 & month == 4 & week == 16 & (minute == 0))

## Plot sub-meter 1, 2 and 3 with title, legend and labels - 60 Minute frequency
plot_ly(houseWeek16, x = ~houseWeek16$DateTime, y = ~houseWeek16$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption April 16Th-22Nd, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```
**Leason:** Depending on the minute we choose, we will see or not some important data points, like the high peaks consumption instants (over 70 watt/hour) shown in the 10 min frequency graph, that are gone in the 60 minutes graph above.

Moving on, let's see a month worth of data, with measurements every 6 hours, choosing an arbitrary minute.
```{r}
## Subset April 2009 - 6 hour frequency
houseWeek16 <- power_df %>%
  filter(year == 2009 & month == 4 & (hour == 0 | hour == 6 | hour == 12 | hour == 18) & (minute == 23))

## Plot sub-meter 1, 2 and 3 with title, legend and labels - 3 hour frequency
plot_ly(houseWeek16, x = ~houseWeek16$DateTime, y = ~houseWeek16$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_3, name = 'Water Heater & AC', mode = 'lines') %>%
 layout(title = "Power Consumption for April, 2009",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```
The Water Heater & AC seem to introduce a lot of noise, let's see only the kitchen and Laundry Room

This way we can tell how many types a week the is someone cooking and/or doing laundry in a month:

```{r}
## Subset the 3rd-4th weeks of April 2009 - 6 hour frequency
houseWeek16 <- power_df %>%
  filter(year == 2009 & month == 4 & (hour == 0 | hour == 6 | hour == 12 | hour == 18) & (minute == 23))

## Plot sub-meter 1, 2 and 3 with title, legend and labels - 3 hour frequency
plot_ly(houseWeek16, x = ~houseWeek16$DateTime, y = ~houseWeek16$Sub_metering_1, name = 'Kitchen', type = 'scatter', mode = 'lines') %>%
 add_trace(y = ~houseWeek16$Sub_metering_2, name = 'Laundry Room', mode = 'lines') %>%
 layout(title = "Power Consumption April, 2009, samples every 6 hours",
 xaxis = list(title = "Time"),
 yaxis = list (title = "Power (watt-hours)"))
```


## Data preparation

Let's take a sample of one observation per week: on Mondays at 8:00pm for 2007, 2008 and 2009, and plot a time series for each sub-meter.
```{r}
## Subset to one observation per week on Mondays at 8:00pm for 2007, 2008 and 2009
house070809weekly <- filter(power_df, weekday == 2 & hour == 20 & minute == 1)
```


### Plot of Sub-meter 3 (Water Heater & AC) time series

```{r}
## Create TS object with SubMeter3
tsSM3_070809weekly <- ts(house070809weekly$Sub_metering_3, frequency=52, start=c(2007,1))

## Plot sub-meter 3 with autoplot (you may need to install these packages)
library(ggplot2)
library(ggfortify)

## Plot sub-meter 3 with autoplot - add labels, color
autoplot(tsSM3_070809weekly, colour = 'green', xlab = "Time", ylab = "Watt Hours", main = "Water Heater & AC", lwd=1)
```


### Plot of Sub-meter 1 (Kitchen) time series

```{r}
## Create TS object with SubMeter1
tsSM1_070809weekly <- ts(house070809weekly$Sub_metering_1, frequency=52, start=c(2007,1))

## Plot sub-meter 1 with autoplot - add labels, color
autoplot(tsSM1_070809weekly, colour = 'blue', xlab = "Time", ylab = "Watt Hours", main = "Kitchen", lwd=1)
```


### Plot of Sub-meter 2 (Laundry Room) time series

```{r}
## Create TS object with SubMeter1
tsSM2_070809weekly <- ts(house070809weekly$Sub_metering_2, frequency=52, start=c(2007,1))

## Plot sub-meter 1 with autoplot - add labels, color
autoplot(tsSM2_070809weekly, colour = 'orange', xlab = "Time", ylab = "Watt Hours", main = "Laundry Room", lwd=1)
```

### Forecasting TS

Now with the available data, we'll try to forecast the power consumption for each sub-meter, using a linear regression model first.

#### Linear model for sub-meter 3:

```{r}
## Apply time series linear regression to the sub-meter 3 ts object and use summary to obtain R2 and RMSE from the model you built
library(forecast)
fitSM3 <- tslm(tsSM3_070809weekly ~ trend + season) 
summary(fitSM3)
```

Now lets create and plot a forecast for sub-meter 3. 
```{r}
## Create the forecast for sub-meter 3. Forecast ahead 20 time periods 
forecastfitSM3 <- forecast(fitSM3, h=20)
## Plot the forecast for sub-meter 3. 
autoplot(forecastfitSM3, xlab = "Year", ylab = "Watt Hours", main = "Water Heater & AC forecast")
```


The graph shows a few things to note:
 - The levels of confidence: 80% and 95% (showing the range at which there is an 85% or 95% probability that the forecasted value will fall into).
 - The levels of confidence shadow show negative values. This is because the confidence level shows a range of acceptance which is equidistant to an edge value greater than, and another one lower than 0.
 - Note that no forecasted values are below 0. 

To correct this situation (of negative values), we can chance the limits of the plot, also we will update the confidence levels:

```{r}
## Create sub-meter 3 forecast with confidence levels 80 and 90
forecastfitSM3c <- forecast(fitSM3, h=20, level=c(80,90))

## Plot sub-meter 3 forecast, limit y and add labels
plot(forecastfitSM3c, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time", main = "Water Heater & AC Forecast")
```

Now let's forecast the other two sub-meters.

#### Linear model for sub-meter 2:
```{r}
## Apply time series linear regression to the sub-meter 2 ts object and use summary to obtain R2 and RMSE from the model you built
fitSM2 <- tslm(tsSM2_070809weekly ~ trend + season) 
summary(fitSM2)
```

Forecast plot for sub-meter 2:

```{r}
## Create sub-meter 2 forecast with confidence levels 80 and 90
forecastfitSM2c <- forecast(fitSM2, h=20, level=c(80,90))

## Plot sub-meter 2 forecast, limit y and add labels
plot(forecastfitSM2c, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time", main = "Laundry Room Forecast")
```

#### Linear model for sub-meter 1:

```{r}
## Apply time series linear regression to the sub-meter 2 ts object and use summary to obtain R2 and RMSE from the model you built
fitSM1 <- tslm(tsSM1_070809weekly ~ trend + season) 
summary(fitSM1)
```


Forecast plot for sub-meter 1:
```{r}
## Create sub-meter 1 forecast with confidence levels 80 and 90
forecastfitSM1c <- forecast(fitSM1, h=20, level=c(80,90))

## Plot sub-meter 1 forecast, limit y and add labels
plot(forecastfitSM1c, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time", main = "Kitchen Forecast")
```

```{r}
plotForecastErrors <- function(forecasterrors)
{
  # make a histogram of the forecast errors:
  mybinsize <- IQR(forecasterrors)/4
  mysd <- sd(forecasterrors)
  mymin <- min(forecasterrors) - mysd*5
  mymax <- max(forecasterrors) + mysd*3
  # generate normally distributed data with mean 0 and standard deviation mysd
  mynorm <- rnorm(10000, mean=0, sd=mysd)
  mymin2 <- min(mynorm)
  mymax2 <- max(mynorm)
  if (mymin2 < mymin) { mymin <- mymin2 }
  if (mymax2 > mymax) { mymax <- mymax2 }
  # make a red histogram of the forecast errors, with the normally distributed data overlaid:
  mybins <- seq(mymin, mymax, mybinsize)
  hist(forecasterrors, col="red", freq=FALSE, breaks=mybins)
  # freq=FALSE ensures the area under the histogram = 1
  # generate normally distributed data with mean 0 and standard deviation mysd
  myhist <- hist(mynorm, plot=FALSE, breaks=mybins)
  # plot the normal curve as a blue line on top of the histogram of forecast errors:
  print(points(myhist$mids, myhist$density, type="l", col="blue", lwd=2))
}
```

### Performance review

All three models perform similarly.

#### Linear models $RSME$ and $R^2$ (residuals) comparison:
```{r}
rsme_list <- c(9.568, 7.403, 8.122)
rsquared_list <- c(0.3375, 0.3006, 0.3544)
model_comparison <- data.frame(rsme_list, rsquared_list)
colnames(model_comparison) <- c("RSME", "RSquared")
rownames(model_comparison) <- c("Kitchen Prediction", "Laundry Room Forecast Prediction", "Water Heater & AC Prediction")
model_comparison
```

Also it seems model's forecast errors do not follow a normal distribution, with center in 0, which indicates the models could be improved further more.

#### Forecast error residuals plot for sub-meter 1
```{r, results = 'hide'}
plotForecastErrors(forecastfitSM1c$residuals) # make a histogram
```

#### Forecast error residuals plot for sub-meter 2
```{r, results = 'hide'}
plotForecastErrors(forecastfitSM2c$residuals) # make a histogram
```

#### Forecast error residuals plot for sub-meter 3
```{r, results = 'hide'}
plotForecastErrors(forecastfitSM3c$residuals) # make a histogram
```


## Seasonal adjustment

### Time series decomposition of sub-meter 1
```{r}
## Decompose Sub-meter 1 into trend, seasonal and remainder
components070809SM1weekly <- decompose(tsSM1_070809weekly)
## Plot decomposed sub-meter 1 
autoplot(components070809SM1weekly, main="Kitchen decomposition of additive time series")
```

```{r}
## Check summary statistics for decomposed sub-meter 1
print("Seasonal")
summary(components070809SM1weekly$seasonal)
print("Trend")
summary(components070809SM1weekly$trend)
print("Random")
summary(components070809SM1weekly$random)
```

### Time series decomposition of sub-meter 2
```{r}
## Decompose Sub-meter 2 into trend, seasonal and remainder
components070809SM2weekly <- decompose(tsSM2_070809weekly)
## Plot decomposed sub-meter 2 
autoplot(components070809SM2weekly, main="Laundry Room decomposition of additive time series")
```

```{r}
## Check summary statistics for decomposed sub-meter 2
print("Seasonal")
summary(components070809SM2weekly$seasonal)
print("Trend")
summary(components070809SM2weekly$trend)
print("Random")
summary(components070809SM2weekly$random)
```

### Time series decomposition of sub-meter 3
```{r}
## Decompose Sub-meter 3 into trend, seasonal and remainder
components070809SM3weekly <- decompose(tsSM3_070809weekly)
## Plot decomposed sub-meter 3 
autoplot(components070809SM3weekly, main = "Water Heater & AC Forecast decomposition of additive time series")
```

```{r}
## Check summary statistics for decomposed sub-meter 3 
print("Seasonal")
summary(components070809SM3weekly$seasonal)
print("Trend")
summary(components070809SM3weekly$trend)
print("Random")
summary(components070809SM3weekly$random)
```


## Holt-Winters Forecasting

### Sub-meter 3 Holt-Winters Forecasting:

```{r}
## Seasonal adjusting sub-meter 3 by subtracting the seasonal component & plot
tsSM3_070809Adjusted <- tsSM3_070809weekly - components070809SM3weekly$seasonal
autoplot(tsSM3_070809Adjusted, main = "Water Heater & AC Forecast Holt-Winters Forecasting")
```

```{r}
## Test Seasonal Adjustment by running Decompose again. Note the very, very small scale for Seasonal
autoplot(decompose(tsSM3_070809Adjusted), main = "Adjusted Water Heater & AC decomposition of additive time series")
```

**"Yes there is a seasonal line, but look at the scale for the seasonal section. -1e-15 through 5e-16. That's a decimal with 15 zeros before 1. A very very small number indeed. For all practical purposes the seasonality has been removed."**


#### HoltWinters Simple Exponential Smoothing
```{r}
## Holt Winters Exponential Smoothing & Plot
tsSM3_HW070809 <- HoltWinters(tsSM3_070809Adjusted, beta=FALSE, gamma=FALSE)
plot(tsSM3_HW070809, ylim = c(0, 25), main="Water Heater & AC Holt-Winters filtering")
```

Perform a new forecast on the data w/o seasonal data:
```{r}
## HoltWinters forecast & plot
tsSM3_HW070809for <- forecast(tsSM3_HW070809, h=25)
plot(tsSM3_HW070809for, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time - Water Heather and AC", main="Water Heater & AC Forecasts from Holt-Winters")
```

```{r}
## Forecast HoltWinters with diminished confidence levels
tsSM3_HW070809forC <- forecast(tsSM3_HW070809, h=25, level=c(10,25))
## Plot only the forecasted area
plot(tsSM3_HW070809forC, ylim = c(0, 9), ylab= "Watt-Hours", xlab="Time - Water Heather and AC", start(2010), main = "Water Heater & AC Holt-Winters forecasts")
```
### TODO: how is this plot different from the forecasted plot from step three of the plan of attack? Which, if any, is more useful? 


### Sub-meter 2 Holt-Winters Forecasting:

```{r}
## Seasonal adjusting sub-meter 2 by subtracting the seasonal component & plot
tsSM2_070809Adjusted <- tsSM2_070809weekly - components070809SM2weekly$seasonal
autoplot(tsSM2_070809Adjusted, main = "Laundry Room Holt-Winters Forecasting")
```

```{r}
## Test Seasonal Adjustment by running Decompose again. Note the very, very small scale for Seasonal
autoplot(decompose(tsSM2_070809Adjusted), main = "Adjusted Laundry Room decomposition of additive time series")
```

#### HoltWinters Simple Exponential Smoothing
```{r}
## Holt Winters Exponential Smoothing & Plot
tsSM2_HW070809 <- HoltWinters(tsSM2_070809Adjusted, beta=FALSE, gamma=FALSE)
plot(tsSM2_HW070809, ylim = c(0, 25), main = "Laundry Room Holt-Winters filtering")
```

Perform a new forecast on the data w/o seasonal data:
```{r}
## HoltWinters forecast & plot
tsSM2_HW070809for <- forecast(tsSM2_HW070809, h=25)
plot(tsSM2_HW070809for, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time", main = "Laundry Room forecasts from Holt-Winters")
```

```{r}
## Forecast HoltWinters with diminished confidence levels
tsSM2_HW070809forC <- forecast(tsSM2_HW070809, h=25, level=c(10,25))
## Plot only the forecasted area
plot(tsSM2_HW070809forC, ylim = c(0, 5), ylab= "Watt-Hours", xlab="Time", start(2010), main = "Laundry Room Holt-Winters forecasts")
```

### Sub-meter 1 Holt-Winters Forecasting:

```{r}
## Seasonal adjusting sub-meter 1 by subtracting the seasonal component & plot
tsSM1_070809Adjusted <- tsSM1_070809weekly - components070809SM1weekly$seasonal
autoplot(tsSM1_070809Adjusted, main = "Kitchen Holt-Winters Forecasting")
```

```{r}
## Test Seasonal Adjustment by running Decompose again. Note the very, very small scale for Seasonal
autoplot(decompose(tsSM1_070809Adjusted), main = "Adjusted Kitchen decomposition of additive time series ")
```

#### HoltWinters Simple Exponential Smoothing
```{r}
## Holt Winters Exponential Smoothing & Plot
tsSM1_HW070809 <- HoltWinters(tsSM1_070809Adjusted, beta=FALSE, gamma=FALSE)
plot(tsSM1_HW070809, ylim = c(0, 25), main = "Kitchen Holt-Winters filtering")
```

Perform a new forecast on the data w/o seasonal data:
```{r}
## HoltWinters forecast & plot
tsSM1_HW070809for <- forecast(tsSM1_HW070809, h=25)
plot(tsSM1_HW070809for, ylim = c(0, 20), ylab= "Watt-Hours", xlab="Time", main = "Kitchen Holt-Winters Forecasts")
```

```{r}
## Forecast HoltWinters with diminished confidence levels
tsSM1_HW070809forC <- forecast(tsSM1_HW070809, h=25, level=c(10,25))
## Plot only the forecasted area
plot(tsSM1_HW070809forC, ylim = c(0, 5), ylab= "Watt-Hours", xlab="Time", start(2010), main = "Kitchen Holt-Winters Forecasts")
```

